---
output: github_document
---

cricinfo
=====================

Functions for Downloading Cricket Data from ESPNCricinfo

Data on all international cricket matches is provided by ESPNCricinfo. This package provides some scraper functions to download the data into tibbles ready for analysis.

Please respect the [ESPNCricinfo terms of use](http://www.espncricinfo.com/ci/content/site/company/terms_use.html).



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(cricinfo)
library(tidyverse)

```


# What does the data look like?

```{r}
data("bowl_inn_data")
bowl_inn_data %>% 
  head() %>% 
  knitr::kable()

```


# Visualising player data

```{r}

ElyssePerry <- fetch_player_data(275487, "T20")
MegLanning <- fetch_player_data(329336, "ODI")
SteveSmith <- fetch_player_data(267192, "Test")
ggplot(MegLanning) + geom_point(aes(x = Date, y = Score, col = NotOut)) +
ggtitle("Meg Lanning ODI Scores")



```

# Visualising the bowling data

```{r}
bowl_inn_data %>% 
  ggplot(aes(x = Wkts, y = Econ)) +
  geom_point(position = "jitter", alpha = 0.3, colour = "steelblue") +
  labs(title = "Wickets vs Economy Rate",
       x = "Wickets",
       y = "Runs per over")
```

# Batting example

```{r}
data(WT20)

bat_data <- cricinfo:::clean_batting_data(WT20) 
bat_data %>% 
  head() %>% 
  knitr::kable()

```

```{r}
bat_data %>% 
  ggplot(aes(x = Country, y = Runs)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90))

```

