---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo=TRUE,
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  fig.path = "man/figures/README-"
)
library(cricketdata)
library(tidyverse)
```

# cricketdata

Functions for Downloading Cricket Data from ESPNCricinfo

Data on all international cricket matches is provided by ESPNCricinfo. This package provides some scraper functions to download the data into tibbles ready for analysis.

Please respect the [ESPNCricinfo terms of use](http://www.espncricinfo.com/ci/content/site/company/terms_use.html).

## Installation

You can install cricketdata from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("ropenscilabs/cricketdata")
```

## Bowling Data

```{r woment20}
# Fetch all Women's T20 data
wt20 <- fetch_cricinfo("T20", "Women", "Bowling")
wt20 %>% head() %>% knitr::kable()

wt20 %>% 
  filter(
    Wickets > 20,
    !is.na(Country)
  ) %>%
  ggplot(aes(y = StrikeRate, x = Country)) +
    geom_boxplot() +
    geom_point(alpha=0.3, col="blue") +
    ggtitle("Women T20: Strike Rates") +
    ylab("Balls per wicket")
```

## Batting Data

```{r menodi}
# Fetch all Australian Men's ODI data by innings
menODI <- fetch_cricinfo("ODI", "Men", "Batting", type="innings", country="Australia")
menODI %>% head() %>% knitr::kable()

menODI %>% 
  ggplot(aes(y = Runs, x = Date)) +
    geom_point(alpha=0.2, col='red') +
    geom_smooth() +
    ggtitle("Australia Men ODI: Runs per Innings")
```

## Fielding Data

```{r indiafielding}
Indfielding <- fetch_cricinfo("Test", "Men", "Fielding", country="India")
Indfielding %>% head() %>% knitr::kable()

Indfielding %>%
  mutate(wktkeeper = (CaughtBehind > 0) | (Stumped > 0)) %>%
  ggplot(aes(x=Matches, y=Dismissals, col=wktkeeper)) +
    geom_point() +
    ggtitle("Indian Men Test Fielding")
```

## Individual player data

```{r meglanning}
MegLanning <- fetch_player_data(329336, "ODI") %>%
  mutate(NotOut = (Dismissal=="not out"))
MLave <- summarise(MegLanning,
    Innings = sum(!is.na(Runs)),
    Average = sum(Runs, na.rm=TRUE) / (Innings - sum(NotOut))
  ) %>%
  pull(Average)
names(MLave) <- paste("Average =", round(MLave,2))

ggplot(MegLanning) + 
  geom_point(aes(x = Start_Date, y = Runs, col = NotOut)) +
  ggtitle("Meg Lanning ODI Scores") +
  geom_hline(aes(yintercept=MLave)) +
  scale_y_continuous(sec.axis = sec_axis(~ ., breaks=MLave))
```

